---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
# Analysis of Fitbit Monitoring Data

## Loading and preprocessing the data
```{r}
mainDir <- getwd()

url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(url, "activity_monitoring_data.zip")
ifelse(!dir.exists(file.path(mainDir, "RepData_PeerAssessment1")),
       dir.create(file.path(mainDir, "RepData_PeerAssessment1")), FALSE)
unzip("activity_monitoring_data.zip", exdir = "./RepData_PeerAssessment1")
setwd("./RepData_PeerAssessment1")

activity <- read.csv("activity.csv")
```
### Here is a histogram showing the total number of steps taken each day
```{r}
daystep <- aggregate(activity$steps~activity$date, FUN=sum, na.rm=TRUE)
names(daystep) <- c("date", "total_steps")
hist(daystep$total_steps, breaks = 10)
```

## What is mean total number of steps taken per day?
```{r}
mean(daystep$total_steps)
median(daystep$total_steps)
```


## What is the average daily activity pattern?
```{r}
avgsteps <- aggregate(activity$steps~activity$interval, FUN=mean, na.rm=TRUE)
names(avgsteps) <- c("interval", "average_steps")
plot(avgsteps$interval, avgsteps$average_steps, type='l', col="blue", xlab="interval", ylab="steps taken", main = "Average steps taken for each 5 min. time interval")
```
### This will show the 5 min. time interval that, on average, contains the maximum number of steps.  It also shows the average step count for that time period.
```{r}
library(plyr)
avg_step_order <- ddply(avgsteps, "average_steps",
          function(x) x[which.max(x$average_steps),])
tail(avg_step_order, 1)
```

## Imputing missing values
### This code finds the number of missing values and imputes a value to replace them.
```{r}
colSums(is.na(activity))
# this shows there are 2,304 missing values in steps column and 0 elsewhere
# now to impute the data using the mice package
library(mice)
tempData <- mice(activity, maxit=50, method="pmm", m = 5, seed = 500)
completeData <- complete(tempData, 1)
#  Show there are no NA values in new data
colSums(is.na(completeData))
```

### This is a histogram of total # of steps taken each day after missing values imputed
```{r}
completesteps <- aggregate(completeData$steps~completeData$date, FUN=sum)
names(completesteps) <- c("date", "total_steps")
hist(completesteps$total_steps, breaks=10)
```

###  Here are the mean and median values after imputing missing values.
```{r}
mean(completesteps$total_steps)
median(completesteps$total_steps)
```
### by comparison we see that the mean is 366.3 lower and median 326 lower after imputing values.


## Are there differences in activity patterns between weekdays and weekends?
###  This is the panel plot comparing the average number of steps taken per 5-minute period across weekdays and weekends.
```{r}
library(timeDate)
library(lattice)
completeData$weekday <- as.factor(isWeekend(completeData$date))
levels(completeData$weekday) <- c("weekday", "weekend")
sortedData <- aggregate(completeData$steps~completeData$interval + 
            completeData$weekday, FUN=mean)
names(sortedData) <- c("interval", "weekday", "avg_steps")
xyplot(avg_steps~interval | factor(weekday), data = sortedData, type='l',
       ylab= "Number of steps", layout=c(1,2))
```
